Accurate inference of microbial differential abundance from taxonomically-biased microbiome measurements

### 2021-09-15 v/6932587bd1ea9c56638c0ebe3ff7885b79b84e8c/_main.tex
Add git head checksum to preface
This commit adds the rendered manuscript associated with
6932587bd1ea9c56638c0ebe3ff7885b79b84e8c.

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\title{Taxonomic bias can create spurious results in microbiome differential-abundance analyses unless it is properly accounted for}
\author{Michael R. McLaren\footnote{North Carolina State University; send correspondence to \href{mailto:m.mclaren42@gmail.com}{\nolinkurl{m.mclaren42@gmail.com}}} \and Karen G. Lloyd\footnote{University of Tennessee} \and Benjamin J. Callahan\footnote{North Carolina State University}}
\date{2021-08-04}

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Taxonomic bias can create spurious results in microbiome differential-abundance analyses unless it is properly accounted for},
  pdfauthor={Michael R. McLaren; Karen G. Lloyd; Benjamin J. Callahan},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[top=1in,bottom=1in,left=1.5in,right=1.5in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{cancel}

@@

@@

Most microbiome research aims to go beyond qualitative descriptions to make quantitative conclusions about associations between specific microbes or community states and key host or environmental properties.
But there is serious disagreement about whether the microbiome measurements made by marker-gene and metagenomic sequencing (MGS) are or can ever be truly quantitative. A primary reason is that taxa vary in how they respond to each step in an MGS protocol, from sample collection to bioinformatic classification.
As a result taxa can differ dramatically (e.g.~10-1000X) in how efficiently they are measured---that is, converted from cells into taxonomically classified sequencing reads---making the relative abundances we observe by MGS inaccurate representations of the actual sample compositions.
Although often associated with variation in primer binding and amplification rates and marker-gene copy-number, large variation in DNA extraction efficiency and in the ability to correctly classify reads among taxa make this taxonomic bias a feature of shotgun metagenomic measurements as much as marker-gene measurements.
Taxonomic bias is protocol specific (\protect\hyperlink{ref-mclaren2019cons}{McLaren, Willis, and Callahan} (\protect\hyperlink{ref-mclaren2019cons}{2019})) and can even vary among batches within an experiment (\protect\hyperlink{ref-yeh2018taxo}{Yeh et al.} (\protect\hyperlink{ref-yeh2018taxo}{2018})); it thus not only makes MGS measurements inaccurate, but also incomparable between studies.
Taxonomic bias thereby poses a major challenge for any seeking to draw quantitative conclusions from MGS studies.

The field's current answer to this challenge is methods standardization, prioritizing consistency over accuracy.
By treating each sample within a study in exactly the same manner, we can ensure that observed differences between samples are not simply due to methodological differences.
Yet standardization does not guarantee that these observed differences reflect the actual differences in taxonomic composition among samples.
Consider what might be the simplest and most common inference made from MGS: How does the proportion of a taxon vary between samples? An observation that \emph{Escherichia} coli makes up 10\% of the sequencing reads in a fecal sample might, thanks to bias, have arisen from a sample in which the proportion of \emph{E. coli} is actually 50\% or 1\%.
Yet many hope that if we observe the proportion of \emph{E. coli} to increase from 10\% to 20\% between two samples, its proportion must have actually doubled---by fixing the taxonomic bias of our measurements through standardization, we have (so the argument goes) fixed the measurement error in both samples so that it will cancel when estimate the fold change.
Unfortunately, the results of \protect\hyperlink{ref-mclaren2019cons}{McLaren, Willis, and Callahan} (\protect\hyperlink{ref-mclaren2019cons}{2019}) show that reality is not so neat.
Because the taxa compete for sequencing effort within a sample, the effect that taxonomic bias has on the observed proportions depends on the taxonomic composition of the sample.
As a result, MGS measurements can lead to spurious inferences about changes across samples even when taxonomic bias is constant for each taxon across samples.

This problem posed by taxonomic bias intersects with a second, growing concern within the microbiome field: that the relative abundances measured by MGS can present a misleading view of absolute-abundance dynamics.
Without making further assumptions (which are often difficult or impossible to verify) one cannot tell whether a doubling in the proportion of a taxon reflects an increase in its abundance or a decrease in the abundances of other taxa.
This concern has motivated the development of various experimental methods to enable researchers to convert the relative abundances measured by MGS into absolute abundances (methods to be reviewed in the Appendix in a future version).
An increasingly popular approach in the study of both environmental and host-associated microbiomes involves making an independent measurement of total microbial abundance in each sample (\protect\hyperlink{ref-props2017abso}{Props et al.} (\protect\hyperlink{ref-props2017abso}{2017}),\protect\hyperlink{ref-kevorkian2018esti}{Kevorkian et al.} (\protect\hyperlink{ref-kevorkian2018esti}{2018}),\protect\hyperlink{ref-lloyd2020evid}{Lloyd et al.} (\protect\hyperlink{ref-lloyd2020evid}{2020}),\protect\hyperlink{ref-tettamantiboshier2020comp}{Tettamanti Boshier et al.} (\protect\hyperlink{ref-tettamantiboshier2020comp}{2020}),\protect\hyperlink{ref-contijoch2019gutm}{Contijoch et al.} (\protect\hyperlink{ref-contijoch2019gutm}{2019}),\protect\hyperlink{ref-vandeputte2017quan}{Vandeputte et al.} (\protect\hyperlink{ref-vandeputte2017quan}{2017}),\protect\hyperlink{ref-vieirasilva2019quan}{Vieira-Silva et al.} (\protect\hyperlink{ref-vieirasilva2019quan}{2019})).
This estimate is then multiplied by the MGS proportions to convert them to estimates of absolute abundance.
This simple calibration procedure directly transfers the error in the MGS proportions due to taxonomic bias to the estimated absolute abundances, leading to estimates that are not only individually inaccurate but can imply inaccurate variation in abundance across samples.

Here we summarize recent advances in our understanding of how taxonomic bias affects our ability to accurately determine how taxa vary in relative and absolute abundance across samples from different conditions---the inference problem known as \emph{differential abundance (DA) analysis}.
We first explain how taxonomic bias can lead to spurious DA results and consider the biological conditions in which such spurious results are likely.
But we then describe several experimental and computational methods that can be used to avoid or correct these errors.
Collectively, these methods may provide practical solutions to mitigating the effects of taxonomic bias on DA analysis for a majority of microbiome studies.

\hypertarget{bias-can-cause-spurious-results-in-proportion-based-da-analysis}{%
\section{Bias can cause spurious results in proportion-based DA analysis}\label{bias-can-cause-spurious-results-in-proportion-based-da-analysis}}

\begin{verbatim}
#> Warning: It is deprecated to specify `guide = FALSE` to remove a guide. Please
#> use `guide = "none"` instead.

#> Warning: It is deprecated to specify `guide = FALSE` to remove a guide. Please
#> use `guide = "none"` instead.
\end{verbatim}

\begin{figure}
\centering
\includegraphics{_main_files/figure-latex/vaginal-example-1.pdf}
\caption{\label{fig:vaginal-example}\textbf{Taxonomic bias can distort differential abundance results even when it is consistent for each taxon across samples.} Panel A shows the actual and observed proportions for hypothetical community samples from two time points, which differ in their relative abundance of three taxa. Panel B shows taxonomic bias in terms of the relative efficiencies of the three taxa against the mean efficiency of each sample; the difference between the taxon's efficiency and the sample's mean (vertical arrows) determines the fold error seen in Panel A. Panel C rearranges the plot from Panel A to show the actual and observed fold changes between time points. The efficiencies of individual taxa were estimated by \protect\hyperlink{ref-mclaren2019cons}{McLaren, Willis, and Callahan} (\protect\hyperlink{ref-mclaren2019cons}{2019}) from mock community data from \protect\hyperlink{ref-brooks2015thet}{Brooks et al.} (\protect\hyperlink{ref-brooks2015thet}{2015}). The abundances are hypothetical but inspired by observations from the human vaginal microbiome; see main text.}
\end{figure}

We illustrate the problems posed by taxonomic bias for differential abundance using an example where the bias of an MGS protocol was explicitly measured using mock communities.
To analyze the effects of taxonomic bias on measurements of the human vaginal microbiome, \protect\hyperlink{ref-brooks2015thet}{Brooks et al.} (\protect\hyperlink{ref-brooks2015thet}{2015}) performed 16S rRNA gene sequencing of mock communities of seven key vaginal species using a similar protocol to that of the Vaginal Human Microbiome Project (VaHMP).
\protect\hyperlink{ref-mclaren2019cons}{McLaren, Willis, and Callahan} (\protect\hyperlink{ref-mclaren2019cons}{2019}) used this data to estimate the relative measurement efficiency of the seven species.
They found that the most efficiently measured species, \emph{Lactobacillus iners}, had an efficiency that was 29X that of the least efficiently measured species, \emph{Gardnerella vaginalis}, primarily due to a greater DNA extraction efficiency and 16S copy number.
A third species, \emph{Atopobium vaginae}, had an efficiency 1.8X that of \emph{G. vaginalis}.
A common aim of human vaginal microbiome studies, including the VaHMP, is to understand the causes and consequences of shifts in communities from being dominated by a particular \emph{Lactobacillus} species to dominance by a non-\emph{Lactobacillus} species such as \emph{G. vaginalis}, which have been associated with deleterious health outcomes including an increased risk of preterm births in pregnant women.
It is therefore important to know whether the efficiency variation found by \protect\hyperlink{ref-mclaren2019cons}{McLaren, Willis, and Callahan} (\protect\hyperlink{ref-mclaren2019cons}{2019}) affects the observed differential abundances between \emph{Lactobacillus}-dominated and \emph{Gardnerella}-dominated communities.

Figure \ref{fig:vaginal-example} illustrates the effects of taxonomic bias on the measured proportions for hypothetical vaginal samples from a single woman at two points in time, during which the community shifts from being dominated by \emph{Lactobacillus} to being dominated by \emph{Gardnerella}.
In each case, we observe a greater proportion of \emph{Lactobacillus} than is truly present, and less of the other two species (Figure \ref{fig:vaginal-example}A).
This positive multiplicative error arises because \emph{Lactobacillus} is more efficiently measured than the average species in each sample.
\protect\hyperlink{ref-mclaren2019cons}{McLaren, Willis, and Callahan} (\protect\hyperlink{ref-mclaren2019cons}{2019}) showed that the observed and actual proportions of a taxon \(i\) in a sample \(s\) are related by the equation
\begin{align}
  \label{eq:error}
  \text{observed}_{i}(s) = \text{actual}_{i}(s) \cdot \frac{\text{efficiency}_{i}}{\text{mean efficiency}(s)}
\end{align}
where \(\text{mean efficiency}(s)=\text{actual}_j(s)*\text{efficiency}_j\) in the denominator is the average efficiency of all taxa in that sample.
Intuitively, Equation \eqref{eq:error} says that the taxon will be overrepresented by the extent to which its measurement efficiency is greater than the average efficiency of cells in the sample.
Figure \ref{fig:vaginal-example}B shows the efficiencies of the three taxa and the mean efficiency in each sample.
The (multiplicative) difference between the taxon efficiencies and the mean efficiency determines the (multiplicative) error in each sample; for example, the mean efficiency in the second sample is 2.5, leading to a 29/2.5=11.6X error in the proportion of \emph{Lactobacillus}.
The error in the second sample is larger---specifically, by 6X---than the first sample because the high proportion of \emph{Gardnerella} has reduced the sample mean efficiency (and thus increased the efficiency of \emph{Lactobacillus} relative to the mean) by 6X.

The enrichment depends on the taxon's efficiency relative to the sample average, rather than the efficiency itself, because the requirement that the proportions of all taxa sum to one locks the taxa in a zero-sum competition.
Notably, competition occurs (though in a less absolute form) even at the level of the raw sequencing reads: DNA extractions and PCR reactions have limited output capacity; high-concentration samples are typically subsampled to fixed, lower concentrations; and sequencing libraries are given limited sequencing effort; so that the experimental process itself typically imposes competition to be sequenced.
And it is competition that prevents the effects of taxonomic bias from canceling when we analyze how the proportions of individual taxa change across samples.

Figure \ref{fig:vaginal-example}C shows the consequences of measurement error on an analysis of the fold change in proportions between the two samples.
In reality, \emph{Gardnerella} changes by 4X, \emph{Lactobacillus} by 0.1X, and \emph{Atopobium} by 0.5X.
Yet what is observed is that \emph{Gardnerella} changes by 32X, \emph{Lactobacillus} by 0.6X, and \emph{Atopobium} by 3X.
In other words, the change in each case is multiplied by 6X, causing a much larger increase in \emph{Gardnerella} and a smaller decrease \emph{Lactobacillus} than actual, and the appearance that \emph{Atopobium} increased when it in fact decreased.
Why do such errors occur?
The multiplicative error in Equation \eqref{eq:error} has two parts: a taxon-specific term in the numerator that is the same in each sample, and a taxon-independent term in the denominator that depends on the composition of the sample through the mean efficiency.
When we compute the fold change in a taxon i between samples s and t,
\begin{align}
  \frac{\text{observed}_{i}(t)}{\text{observed}_{i}(s)}=\frac{\text{actual}_{i}(t)}{\text{actual}_{i}(s)}*\frac{\text{efficiency}_{i}}{\text{efficiency}_{i}}*\frac{\text{mean efficiency}(s)}{\text{mean efficiency}(t)}
\end{align}
the constant efficiency in the numerator of Equation \eqref{eq:error} cancels, but the varying mean efficiency does not, leaving an error that is equal to the inverse change in mean efficiency.
As a result, we are here left inferring that the fold-changes in proportions have all increased less (or decreased more) than they truly have.
In our example, the mean efficiency decreased by 6X (from 15 to 2.5) due to the shift in dominant taxon from the high-efficiency \emph{Lactobacillus} to the low-efficiency \emph{Gardnerella}, causing the observed fold changes to increase by 6X above the actual.
This increase distorted the magnitudes of the changes in \emph{Lactobacillus} and \emph{Gardnerella} (decreasing the first and increasing the second) and changed the sign of \emph{Atopobium} from negative to positive.
Thus partial cancelling in the effect of a constant multiplicative bias may not be sufficient to protect against spurious results.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\title{Accurate inference of microbial differential abundance from taxonomically-biased microbiome measurements}
\author{Michael R. McLaren\footnote{North Carolina State University; send correspondence to \href{mailto:m.mclaren42@gmail.com}{\nolinkurl{m.mclaren42@gmail.com}}} \and Karen G. Lloyd\footnote{University of Tennessee} \and Benjamin J. Callahan\footnote{North Carolina State University}}
\date{2021-09-15}

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Accurate inference of microbial differential abundance from taxonomically-biased microbiome measurements},
  pdfauthor={Michael R. McLaren; Karen G. Lloyd; Benjamin J. Callahan},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[top=1in,bottom=1in,left=1.5in,right=1.5in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{cancel}

@@

\emph{This manuscript was rendered from commit 6932587bd1ea9c56638c0ebe3ff7885b79b84e8c.}

@@

The most basic question we can ask about microbial communities after which taxa are found in an ecosystem is, how do the abundances of these taxa vary in their abundance---across space, time, and host or environmental conditions?
Advances in sequencing technology allow us to now simultaneously measure the abundances of 100s to 1000s of species using marker-gene and shotgun-metagenomic sequencing (jointly, MGS).
Although standard MGS measurements lose information about total microbial density---and so are typically used to analyze the abundances of taxa relative to each other or their total---new studies are increasingly employing strategies to enable the analysis of cell density or other measures of ``absolute abundance.''
These relative and absolute abundances serve as the basis for a \emph{differential-abundance (DA) analysis}, in which the change in abundance of a microbial taxon across samples or conditions is used to learn about the biology of the taxon and its impact on the host and other microbes as well as to detect predictive biomarkers of host and environmental health and disease.

Although MGS-based DA analysis has been widely deployed and achieved many notable successes, it faces serious concerns over accuracy and reproducibility due to the inherent technical limitations of MGS measurements.
In particular, MGS measurements are \emph{taxonomically biased}: Taxa vary dramatically (e.g.~10-1000X) in how efficiently they are measured---that is, converted from cells into taxonomically classified sequencing reads---by a given MGS protocol.
As a result, the abundance measurements obtained by MGS are inaccurate representations of the actual abundances and also tend to differ across protocols (\protect\hyperlink{ref-mclaren2019cons}{McLaren, Willis, and Callahan} (\protect\hyperlink{ref-mclaren2019cons}{2019})), studies, and even experimental batches (\protect\hyperlink{ref-yeh2018taxo}{Yeh et al.} (\protect\hyperlink{ref-yeh2018taxo}{2018})).
This bias arises from variation in how taxa respond to each step in an MGS protocol, from sample collection to bioinformatic classification.
Although often associated with variation in primer binding and amplification rates and marker-gene copy-number, large variation in DNA extraction efficiency and in the ability to correctly classify reads make taxonomic bias a feature of both shotgun and marker-gene measurements.
The error it causes have been found to in some cases to supersede sizable biological differences (e.g. \protect\hyperlink{ref-lozupone2013meta}{Lozupone et al.} (\protect\hyperlink{ref-lozupone2013meta}{2013})) and has plausibly caused replication failures for prominent findings such as the association of decreased Bacteroides and increased Firmicutes in stool with obesity (\protect\hyperlink{ref-finucane2014atax}{Finucane et al.} (\protect\hyperlink{ref-finucane2014atax}{2014})) and the association of certain taxa in the vagina of pregnant women with preterm birth (\protect\hyperlink{ref-callahan2017repl}{Callahan et al.} (\protect\hyperlink{ref-callahan2017repl}{2017})).

The typical approach to countering taxonomic bias in DA analysis is to standardize the measurement protocol used within a given study.
In broad strokes, the thinking is that the measurements of samples measured by the same protocol will be affected by bias in the same way and so the inferred differences between samples (the focus of DA analysis) will be unaffected.
For example, if taxonomic bias consistently causes the measured proportion of a given species to be 10X too high, we can still accurately infer its fold changes across samples.
\protect\hyperlink{ref-kevorkian2018esti}{Kevorkian et al.} (\protect\hyperlink{ref-kevorkian2018esti}{2018}) and \protect\hyperlink{ref-lloyd2020evid}{Lloyd et al.} (\protect\hyperlink{ref-lloyd2020evid}{2020}) were the first (to our knowledge) to make this argument explicit mathematically; however, \protect\hyperlink{ref-mclaren2019cons}{McLaren, Willis, and Callahan} (\protect\hyperlink{ref-mclaren2019cons}{2019}) used a theoretical model (supported by measurements of defined bacterial communities) to show that consistent taxonomic bias can lead to variable fold errors in measured proportions.
These varying errors can lead to spurious conclusions for how the proportion of a taxon varies across samples, even in the direction of change (for example, causing a taxon that decreases appear to increase).
Yet \protect\hyperlink{ref-mclaren2019cons}{McLaren, Willis, and Callahan} (\protect\hyperlink{ref-mclaren2019cons}{2019}) also found that certain changes in relative abundance---in particular, fold changes in the ratios among species---are robust to the effects of bias.
The implications of these findings for changes in absolute abundance---which remain subject to taxonomic bias in the underlying MGS measurement---and for DA analysis across many species and many samples---as commonly done in microbiome association testing---have yet to be investigated.

Here we use a combination of theoretical analysis, simulation, and re-analysis of published experiments to consider when and why taxonomic bias in MGS measurements leads to spurious results in DA analysis of relative and absolute abundance.
Our analysis clarifies how the folk wisdom that taxonomic bias does not affect the analysis of change across samples is only partially correct and can give a false sense of security in the accuracy of DA results.
Yet we also present several potential solutions---methods for quantifying, correcting, or otherwise accounting for the effect of taxonomic bias in DA analyses that can be deployed today with only modest changes to existing experimental and analytical workflows.
Over time, application of these methods to past and future experiments will provide crucial quantitative information about taxonomic bias and the conditions under which spurious results arise for various DA methodologies.
Collectively, these methods and insights may provide practical solutions to taxonomic bias in DA analysis and the confidence that is necessary to codify the statistical findings of microbiome studies into readily-translatable scientific knowledge.

\hypertarget{abundance-measurement}{%
\section{How taxonomic bias affects abundance measurements}\label{abundance-measurement}}

To understand how bias affects the measured differential abundance between samples, we first consider how it affects measurements of individual samples.

Our primary tool for understanding the impact of taxonomic bias MGS measurement is the theoretical model of MGS measurement developed and empirically validated by \protect\hyperlink{ref-mclaren2019cons}{McLaren, Willis, and Callahan} (\protect\hyperlink{ref-mclaren2019cons}{2019}).
This model is the simplest model of MGS measurement that includes multiplicative taxonomic bias while respecting the \emph{compositional} nature of sequencing measurements, in which the total read count for a sample is unrelated to its total cell number or density (\protect\hyperlink{ref-gloor2017micr}{Gloor et al.} (\protect\hyperlink{ref-gloor2017micr}{2017})).
In this model, we suppose there are a set of \emph{atomic taxa}, that differ in how efficiently they are measured but within which all cells behave similarly.
For concreteness, we equate atomic taxa with species.
We ignore the possibility of contamination or false-positive taxonomic assignments---reads assigned to a given species and sample really do come from that species and sample.

Our model stipulates that the assigned read count of species \(i\) in sample \(a\) equal equals its cell density multiplied by a species-specific, sample-independent factor and a species-independent, sample-specific factor,
\begin{align}
  \label{eq:measurement-model}
  \text{reads}_{i}(a)
  = \text{density}_{i}(a) \quad \cdot
    \underbrace{\text{efficiency}_{i}}_{\substack{\text{species specific,} \\  \text{sample independent}}}
    \cdot \quad
    \underbrace{\text{sequencing effort}(a)}_{\substack{\text{species independent,} \\  \text{sample specific}}}.
\end{align}
The species-specific factor is the \emph{relative measurement efficiency} (or \emph{efficiency} for short) of the species, which represents how much more easily that species is measured (converted from cells to assigned reads) relative to an abitrarily chosen reference species (\protect\hyperlink{ref-mclaren2019cons}{McLaren, Willis, and Callahan} (\protect\hyperlink{ref-mclaren2019cons}{2019})).
The sample-specific factor, which we call the \emph{sequencing effort} of the sample, reflects the fact that the number of reads per unit cell density varies among samples even in the absence of taxonomic bias due to variation in total density, library normalization, and total sequencing-run output.
Equation \eqref{eq:measurement-model} implies that the total reads from all species in the sample equal
\begin{align}
  \label{eq:total-reads}
  \text{total reads}(a)
    = \text{total density}(a) \cdot \text{mean efficiency}(a) \cdot \text{sequencing effort}(a),
\end{align}
where
\begin{align}
  \label{eq:mean-efficiency}
  \text{mean efficiency}(a) 
  \equiv \frac{\sum_{j}\text{density}_j(a)\cdot \text{efficiency}_j}{\text{total density}(a)}
\end{align}
is the average or mean efficiency of cells in the sample.

\hypertarget{relative-abundances-proportions-and-ratios}{%
\subsection{Relative abundances (proportions and ratios)}\label{relative-abundances-proportions-and-ratios}}

The term \emph{relative abundance} carries multiple meanings in microbiome science.
Microbiome researchers often use the relative abundance of a species in a sample to refer to its \emph{proportion}, or number of cells from that species divided by the total number of cells from the taxonomic domain of interest (e.g.~Prokaryotes).
A second view, drawn from the statistical framework of Compositional Data Analysis, equates the relative abundances of species within a sample with the \emph{ratios} of cell densities among two or more species (e.g.~a ratio of 5:4:2 among three species).
By this view, it is meaningless to refer to the relative abundance of an individual species.
Although other interpretations exist\footnote{e.g.~the rank order within a sample, or value associated with a species after a centered-log-ratio transform}, we focus on the proportion and ratio views as they underly most DA analysis methods (and microbiome analysis in general), for both relative and absolute abundances, and the differences between them help to clarify the issue over when and why the effects of bias cancel in an analysis.

The standard estimate of the proportion of species \(i\) in sample \(a\) given by the ratio of its read count to the total,
\begin{align*}
  \hat{\text{prop}}_{i}(a) = \frac{\text{reads}_i(a)}{\text{total reads(a)}}.
\end{align*}
The effect of taxonomic bias on this estimate is to create a multipliative error equal to the ratio of the species' efficiency to the mean efficiency in the sample,
\begin{align}
  \label{eq:prop-error}
  \hat{\text{prop}}_{i}(a)
  &= \text{prop}_{i}(a) \cdot \frac{\text{efficiency}_{i}}{\text{mean efficiency}(a)}.
\end{align}
Intuitively, Equation \eqref{eq:prop-error} says that a species is measured as having a larger-than-actual proportion in direct relation to how much more more efficiently its cells are measured than the average cell in the sample (Figure \ref{fig:error-proportions}).
Consequently, the same species can be over or under measured in terms of its proportion.
For instance, in samples from two hypothetical communities in Figure \ref{fig:error-proportions}, Species 3 has an efficiency of 6 and is under-estimated in Sample 1 (which has a mean efficiency of 8.33) but over-estimated in Sample 2 (which has a mean efficiency of 3.15).

The estimated ratio between species \(i\) and \(j\), given by the ratio of read counts,
\begin{align*}
  \hat{\text{ratio}}_{i/j}(a) = \frac{\text{reads}_i(a)}{\text{reads}_j(a)},
\end{align*}
has a multiplicative error equal to the ratio in the species' efficiencies,
\begin{align}
  \label{eq:ratio-error}
  \hat{\text{ratio}}_{i/j}(a)
%  \frac{\text{reads}_{i}(a)}{\text{reads}_{j}(a)}
  &= \frac{\text{density}_{i}(a)}{\text{density}_{j}(a)} \cdot \frac{\text{efficiency}_{i}}{\text{efficiency}_{j}}.
\end{align}
Equation \eqref{eq:ratio-error} says that the ratio between two species is over-estimated in direct relation to how much more efficiently cells of the first species is measured relative to the second.
In contrast to estimates of proportions, estimates of species ratios have a consistent multiplicative error across samples.
For instance, in the example of Figure \ref{fig:error-proportions}, the ratio of Species 3 (with an efficiency of 6) to Species 1 (with an efficiency of 1) is over-estimated by a factor of 6 in both communities despite their varying compositions.

\begin{figure}
\includegraphics[width=0.9\linewidth]{/tmp/bookdown-differential-abundance-theory/error-proportions} \caption{\textbf{Taxonomic bias creates sample-dependent multiplicative errors in species proportions, which can lead to inaccurate fold changes between samples.} Top row: Error in proportions measured by MGS in two microbiome samples that contain different relative abundances of three species. Bottom row: Error in the estimated fold-change in the third species that is derived from these measurements.}\label{fig:error-proportions}
\end{figure}

\hypertarget{absolute-abundances-cell-densities}{%
\subsection{Absolute abundances (cell densities)}\label{absolute-abundances-cell-densities}}

Various experimental and computational methods have or can be used to supplement MGS measurements so as to convert the relative-abundance information in the read counts into estimates of (absolute) cell densities;
however, these methods can largely be partitioned conceptually into those that use the estimated proportions or the estimated ratios provided by MGS.
The error in density estimates due to taxonomic bias therefore follows that of the proportion and ratio estimates described above.
